---
title: "Bacillus growth curves"
author: "Jasmine Ahmed and Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin = 2.54cm
---

# Bacillus subtilis wild type and mutant in DSM: OD, heat, and CFU

# Clear workspace and set directory

```{r setup}
rm(list = ls())
setwd("~/GitHub/symcel/")
getwd()
```

## Load packages and functions

```{r}
require("png")
require("dplyr")
library("tidyverse")
require("grid")
require("tibble")
require("knitr")
require("extrafont")
require("ggrepel");
require("gridExtra")
require("contrast")
sem <- function(x) sqrt(var(x)/length(x))
cv <- function(x) (sd(x)/mean(x))*100
```

# Load OD data

```{r}
data.od <- read.csv("data/20240715_OD.csv")
design.od <- read.csv("data/20240715_OD_design.csv")
```

# Change time for OD data

```{r}
# Change times to numeric (hours)
time <- as.numeric(sub("^(\\d+):(\\d+).*", "\\1.\\2", data.od$Time))
unfilt.data.od <- data.frame(time, data.od[, 2:97])
start_time <- 1
end_time <- 24
data.od <- unfilt.data.od %>%
  filter(time >= start_time & time <= end_time)
```

# Create look-up table for OD

```{r}
# This code chunk reads in the experimental design file
# Converts 96-well plate layout into long form
# Then creates column number reference number

# Subset the design file
treat.od <- design.od[1:8,1:12]
wells.od <- design.od[9:16,1:12]

# convert 96 well plate format to long format
treat.long.od <- as.data.frame(pivot_longer(treat.od, cols = everything(), 
              names_to = "Treatment", values_to = "Value"))
wells.long.od <- as.data.frame(pivot_longer(wells.od, cols = everything(), 
              names_to = "Treatment", values_to = "Value"))

# Create vector of column numbers
col.vect.od <- seq(1, 96)

# Combine into one look up table
look.table.od <- data.frame(treat.long.od[,2], wells.long.od[,2], col.vect.od)
colnames(look.table.od) <- c("treat", "well", "col")
```


# Pull out wells by treatment and make OD data frames

```{r}
# Following code pulls out vector of column numbers corresponding to a medium
# Then creates a data frame with time and ODs
# Need to add one (+1) because of how time in read in
# Note: there are some wells with no growth that could be removed here

# Blanks (empty)
blank.col <- look.table.od %>%
  filter(treat == "none") %>%
  pull(col)
blank <- data.frame(data.od[,1], data.od[, blank.col+1])
colnames(blank)[1] <- "time"

# PBS (no cell, no media control)
pbs.col <- look.table.od %>%
  filter(treat == "pbs") %>%
  pull(col)
pbs <- data.frame(data.od[,1], data.od[, pbs.col+1])
colnames(pbs)[1] <- "time"

# DSM (no cell, media control)
dsm.col <- look.table.od %>%
  filter(treat == "dsm") %>%
  pull(col)
dsm <- data.frame(data.od[,1], data.od[, dsm.col+1])
colnames(dsm)[1] <- "time"

# B. subtilis sporulating wild type
wt.col <- look.table.od %>%
  filter(treat == "delta") %>%
  pull(col)
wt.od <- data.frame(data.od[,1], data.od[, wt.col+1])
colnames(wt.od)[1] <- "time"

# B. subtilis non-sporulating mutant
mut.col <- look.table.od %>%
  filter(treat == "spo") %>%
  pull(col)
mut.od <- data.frame(data.od[,1], data.od[, mut.col+1])
colnames(mut.od)[1] <- "time"
```

# Make OD figure

```{r}
# Following line initiate location to write file
png(filename ="figures/Growth_20240716.png", width = 1200, height = 1200, res = 96 * 2) 

# Start plot 
plot.new()

# Define plotting margins
par(mar = c(7, 7, 5, 7))

# Make plot for the first wild type sporulator
plot(wt.od[,1], wt.od[,2], xlim = c(0, 25), 
     ylim = c(0, 1.5), type = "l", 
     lty = 1, col = "blue", lwd = 2, ylab = "", xlab = "", 
     cex.lab = 1.5, las = 1, yaxt = "n", xaxt = "n")

# Add lines for other blank replicates (color = grey)
# We just plotted column 2 and 1
# Now we plot column 3 and 1, and so forth
for(i in 3:ncol(wt.od)) {
  lines(wt.od[,1], wt.od[,i], col = "blue")  
}

# Add non-sporulator lines
# this time plot column 2 and 1, and so forth
for(i in 2:ncol(mut.od)) {
  lines(mut.od[,1], mut.od[,i], col = "red")  
}

# Add box around plot
box(lwd = 2)

# Add axis ticks and labels
axis(side = 2, lwd.ticks = 2, las = 1, cex.axis = 1.25, 
     labels = c("0", "0.5", "1.0", "1.5"), at = c(0, 0.5, 1, 1.5))

axis(side = 4, labels = F, lwd.ticks = 2, at = c(0, 0.5, 1, 1.5))

axis(side = 1, lwd.ticks = 2, cex.axis = 1.25, las = 1, mgp = c(1, 1, 0),
    labels = c("0", "5", "10", "15", "20", "25"), at = c(0, 5, 10, 15, 20, 25))

axis(side = 3, labels = F, lwd.ticks = 2, las = 1, cex.axis = 1, 
   at = c(0, 5, 10, 15, 25))


# Add axis labels
mtext('Biomass (OD600)', side = 2, outer = TRUE, cex = 1.5, 
      line = -3.5, adj = 0.55)

mtext('Time (h)', side = 1, outer = TRUE, cex = 1.5, 
      line = -4, adj = 0.5)

# Add a legend
legend("topright", legend = c("sporulator", "non-sporulator"), col = c("blue", "red"), lty = 1, 
       cex = 1.5, bty = "n", bg = "transparent", seg.len = 1.25, lwd = 1.25)


# Close plot device (completes writing of file)
dev.off()
graphics.off()

# Shows plot in R console window
img <- readPNG("figures/Growth_20240716.png")
grid.raster(img)
```


# Load heat data

```{r}
data.ht <- read.csv("data/20240715_symcel.csv")
design.ht <- read.csv("data/20240715_symcelll_design.csv")
```

# Change time for heat data

```{r}
# Change times to numeric (hours)
time.s <- as.numeric(sub("^(\\d+):(\\d+).*", "\\1.\\2", data.ht[,1]))
colnames(data.ht)[colnames(data.ht) == "Time..s."] <- "time"
time.h <- data.ht[,1]/60/60
unfilt.data.ht <- data.frame(time.h, data.ht[,2:48])
# Filter time (remove the acclimation period)
start_time <- 1
end_time <- 25
data.ht <- unfilt.data.ht %>%
  filter(time.h >= start_time & time.h <= end_time)
```

# Create look-up table for heat

```{r}
# This code chunk reads in the experimental design file
# Converts 96-well plate layout into long form
# Then creates column number reference number

# Subset the design file
treat.ht <- design.ht[1:6,1:8]
wells.ht <- design.ht[7:12,1:8]

# convert 96 well plate format to long format
treat.long.ht <- as.data.frame(pivot_longer(treat.ht, cols = everything(), 
              names_to = "Treatment", values_to = "Value"))
wells.long.ht <- as.data.frame(pivot_longer(wells.ht, cols = everything(), 
              names_to = "Treatment", values_to = "Value"))

# Create vector of column numbers
col.vect.ht <- seq(1, 48)

# Combine into one look up table
look.table.ht <- data.frame(treat.long.ht[,2], wells.long.ht[,2], col.vect.ht)
colnames(look.table.ht) <- c("treat", "well", "col")
```


# Pull out wells by treatment and make data frames for heat

```{r}
# Following code pulls out vector of column numbers corresponding to a medium
# Then creates a data frame with time and ODs
# Need to add one (+1) because of how time in read in
# Note: there are some wells with no growth that could be removed here

# Reference cells
#ref.col <- look.table.ht %>%
#  filter(treat == "reference") %>%
#  pull(col)
#ref <- data.frame(data.ht[,1], data.ht[, ref.col+1])
#colnames(ref)[1] <- "time"

# B. subtilis sporulating wild type
wt.col <- look.table.ht %>%
  filter(treat == "delta") %>%
  pull(col)

# Remove wells with no growth
# B8 (16)
# C8 (24)
# B7 (15)
# C7 (23)

wt.remove <- c(15, 16, 23, 24)
wt.col <- wt.col[!wt.col %in% wt.remove]

wt.ht <- data.frame(data.ht[,1], data.ht[, wt.col+1])
colnames(wt.ht)[1] <- "time"

# B. subtilis non-sporulating mutant
mut.col <- look.table.ht %>%
  filter(treat == "spo") %>%
  pull(col)

# Remove wells with no growth
# D7 (31)
# D8 (32)
# E7 (39)
# E8 (40)

mut.remove <- c(31, 32, 39, 40)
mut.col <- mut.col[!mut.col %in% mut.remove]

mut.ht <- data.frame(data.ht[,1], data.ht[, mut.col+1])
colnames(mut.ht)[1] <- "time"
```

# Make heat figure

```{r}
# Following line initiate location to write file
png(filename ="figures/Heat_20240716.png", 
    width = 1200, height = 1200, res = 96 * 2) 

# Start plot 
plot.new()

# Define plotting margins
par(mar = c(7, 7, 5, 7))

# Make plot for the first wild type sporulator
plot(wt.ht[,1], wt.ht[,2], xlim = c(0, 25), 
     ylim = c(-30, 50), type = "l", 
     lty = 1, col = "blue", lwd = 2, ylab = "", xlab = "", 
     cex.lab = 1.5, las = 1, yaxt = "n", xaxt = "n")

# Add lines for other blank replicates (color = grey)
# We just plotted column 2 and 1
# Now we plot column 3 and 1, and so forth
for(i in 3:ncol(wt.ht)) {
  lines(wt.ht[,1], wt.ht[,i], col = "blue")  
}

# Add non-sporulator lines
# this time plot column 2 and 1, and so forth
for(i in 2:ncol(mut.ht)) {
  lines(mut.ht[,1], mut.ht[,i], col = "red")  
}

# Add box around plot
box(lwd = 2)

# Add axis ticks and labels
axis(side = 2, lwd.ticks = 2, las = 1, cex.axis = 1.25, 
     labels = c("-25", "0", "25", "50"), at = c(-25, 0, 25, 50))

axis(side = 4, labels = F, lwd.ticks = 2, at = c(-25, 0, 25, 50))

axis(side = 1, lwd.ticks = 2, cex.axis = 1.25, las = 1, mgp = c(1, 1, 0),
    labels = c("0", "5", "10", "15", "20", "25"), at = c(0, 5, 10, 15, 20, 25))

axis(side = 3, labels = F, lwd.ticks = 2, las = 1, cex.axis = 1, 
   at = c(0, 5, 10, 15, 25))


# Add axis labels
mtext('Heat (μW)', side = 2, outer = TRUE, cex = 1.5, 
      line = -3.5, adj = 0.55)

mtext('Time (h)', side = 1, outer = TRUE, cex = 1.5, 
      line = -4, adj = 0.5)

# Add a legend
legend("topright", legend = c("sporulator", "non-sporulator"), col = c("blue", "red"), lty = 1, 
       cex = 1.5, bty = "n", bg = "transparent", seg.len = 1.25, lwd = 1.5)


# Close plot device (completes writing of file)
dev.off()
graphics.off()

# Shows plot in R console window
img <- readPNG("figures/Heat_20240716.png")
grid.raster(img)
```

# Load CFU data

```{r}
cfu.raw <- read.csv("data/20240725_CFU.csv")
colnames(cfu.raw)[1] <- "t"
```

# Calculate CFU/mL
```{r}
cfu.a <- cfu.raw %>%
  mutate(cfu.ml.4 = ((cfu.raw$cfu.4/cfu.raw$vol) * 1000)/ 10^-4)

cfu <- cfu.a %>%
  mutate(cfu.ml.5 = ((cfu.raw$cfu.5/cfu.raw$vol) * 1000)/ 10^-5)
```

# Are there any spores in the mutant strain?
```{r}
mut.spore <- cfu %>% filter(strain == "mutant", heat == "yes") %>%
  summarise(sum.10_4 = sum(cfu.ml.4), sum.10_5 = sum(cfu.ml.4))

# Answer = no spores recovered after heat treatment
```

# Data for time series
```{r}
# wild type vegetative cells
wt.veg <- cfu %>%
  filter(strain == "wt", heat == "no") %>%
  select(inc.time, rep, cfu.ml.4, cfu.ml.5) %>%
  mutate(mean.cfu = rowMeans(select(., cfu.ml.4, cfu.ml.5)))

wt.veg.sum <- wt.veg %>%
  group_by(inc.time) %>%
  summarise(
    mean.cfu = mean(mean.cfu, na.rm = TRUE),
    count = n(),  # Count the number of observations per group
    sem.cfu = ifelse(count > 1, sd(c(mean.cfu, na.rm = TRUE)) / sqrt(count), NA)
  )

# wild type spores
wt.spore <- cfu %>%
  filter(strain == "wt", heat == "yes") %>%
  select(inc.time, rep, cfu.ml.4, cfu.ml.5) %>%
  mutate(mean.cfu = rowMeans(select(., cfu.ml.4, cfu.ml.5)))

wt.spore.sum <- wt.spore %>%
  group_by(inc.time) %>%
  summarise(
    mean.cfu = mean(mean.cfu, na.rm = TRUE),
    count = n(),  # Count the number of observations per group
    sem.cfu = ifelse(count > 1, sd(c(mean.cfu, na.rm = TRUE)) / sqrt(count), NA)
  )

# Filter out points with mean.cfu of 0
wt.spore.filt <- wt.spore.sum %>%
  filter(mean.cfu > 0)

# mutant vegetative cells
mut.veg <- cfu %>%
  filter(strain == "mutant", heat == "no") %>%
  select(inc.time, rep, cfu.ml.4, cfu.ml.5) %>%
  mutate(mean.cfu = rowMeans(select(., cfu.ml.4, cfu.ml.5)))

mut.veg.sum <- mut.veg %>%
  group_by(inc.time) %>%
  summarise(
    mean.cfu = mean(mean.cfu, na.rm = TRUE),
    count = n(),  # Count the number of observations per group
    sem.cfu = ifelse(count > 1, sd(c(mean.cfu, na.rm = TRUE)) / sqrt(count), NA)
  )


# sporulation efficiency

spore.efficiency <- wt.spore$mean.cfu[wt.spore$mean.cfu != 0] /
        wt.veg$mean.cfu[wt.spore$mean.cfu != 0]

mean.spore.efficiency <- mean(spore.efficiency)

```

# Make heat figure

```{r}
# Following line initiate location to write file
png(filename ="figures/CFU_20240725.png", 
    width = 1200, height = 1200, res = 96 * 2) 

# Start plot 
plot.new()

# Define plotting margins
par(mar = c(7, 7, 5, 7))

# Make plot for wild type vegetative cells
plot(wt.veg.sum$inc.time, wt.veg.sum$mean.cfu, xlim = c(-2, 32), 
     ylim = c(1e7, 1e9), type = "l", 
     lty = 1, col = "blue", lwd = 2, ylab = "", xlab = "", 
     cex.lab = 1.5, las = 1, yaxt = "n", xaxt = "n", log = "y")

# Add points (symbols) on the line
points(wt.veg.sum$inc.time, wt.veg.sum$mean.cfu, pch = 16, col = "blue") 
#arrows(x0 = wt.veg.sum$inc.time, y0 = wt.veg.sum$mean.cfu, 
#       y1 = wt.veg.sum$mean.cfu + wt.veg.sum$sem.cfu, 
#       angle = 90, length = 0.1, lwd = 2, col = "blue")
#arrows(x0 = wt.veg.sum$inc.time, y0 = wt.veg.sum$mean.cfu, 
#       y1 = wt.veg.sum$mean.cfu - wt.veg.sum$sem.cfu, 
#       angle = 90, length = 0.1, lwd = 2, col = "blue")

# Make plot for wild type spores
lines(wt.spore.filt$inc.time, wt.spore.filt$mean.cfu, 
      col = "lightblue", lty = 1, lwd = 2)
points(wt.spore.filt$inc.time, wt.spore.filt$mean.cfu, pch = 16, col = "lightblue") 
#arrows(x0 = wt.spore.sum$inc.time, y0 = wt.spore.sum$mean.cfu, 
#       y1 = wt.spore.sum$mean.cfu + wt.spore.sum$sem.cfu, 
#       angle = 90, length = 0.1, lwd = 2, col = "lightblue")
#arrows(x0 = wt.spore.sum$inc.time, y0 = wt.spore.sum$mean.cfu, 
#       y1 = wt.spore.sum$mean.cfu - wt.spore.sum$sem.cfu, 
#       angle = 90, length = 0.1, lwd = 2, col = "lightblue")

# Make plot for wild type mutant
lines(mut.veg.sum$inc.time, mut.veg.sum$mean.cfu, 
      col = "red", lty = 1, lwd = 2)
points(mut.veg.sum$inc.time, mut.veg.sum$mean.cfu, pch = 16, col = "red") 
#arrows(x0 = mut.veg.sum$inc.time, y0 = mut.veg.sum$mean.cfu, 
#       y1 = mut.veg.sum$mean.cfu + mut.veg.sum$sem.cfu, 
#       angle = 90, length = 0.1, lwd = 2, col = "red")
#arrows(x0 = mut.veg.sum$inc.time, y0 = mut.veg.sum$mean.cfu, 
#       y1 = mut.veg.sum$mean.cfu - mut.veg.sum$sem.cfu, 
#       angle = 90, length = 0.1, lwd = 2, col = "red")

# Add box around plot
box(lwd = 2)

# Add axis ticks and labels
axis(2, at = c(1e7, 1e8, 1e9), cex.axis = 1.25,
      labels = expression(10^7, 10^8, 10^9), las = 1)

axis(4, labels = F, lwd.ticks = 2, las = 1, cex.axis = 1.25,
      at = c(10^7, 10^8, 10^9))

axis(side = 1, lwd.ticks = 2, cex.axis = 1.25, las = 1, mgp = c(1, 1, 0),
    labels = c("0", "10", "20", "30"), at = c(0, 10, 20, 30))

axis(side = 3, labels = F, lwd.ticks = 2, las = 1, cex.axis = 1, 
   at = c(0, 10, 20, 30))


# Add axis labels
mtext('CFU (#/mL)', side = 2, outer = TRUE, cex = 1.5, 
      line = -3.5, adj = 0.55)

mtext('Time (h)', side = 1, outer = TRUE, cex = 1.5, 
      line = -4, adj = 0.5)

# Add a legend
legend("topleft", legend = c("wild type: vegetative", 
      "wildtype: spore", "mutant = vegetative"), 
       col = c("blue","lightblue", "red"), lty = 1, 
       cex = 1, bty = "n", bg = "transparent", seg.len = 1.25, lwd = 1.5)


# Close plot device (completes writing of file)
dev.off()
graphics.off()

# Shows plot in R console window
img <- readPNG("figures/CFU_20240725.png")
grid.raster(img)
```



